### 2017-11-14 submission/manuscript.Rmd
add Will's comments discussion up to P2

Our study identifies clear diversity changes both at the community level and within individual genera that are present in individuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC associated genera increases the relative risk of carcinoma it does not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community that contribute to the formation of an adenoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma. The drasitically increased relative risk of disease, when considering the CRC associated genera, is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity. It is also possible that in a driver-passenger scenario that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on relative risk for increasing positivity. Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.

>>>>

Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions. Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma. This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma. Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included. CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions. These observations suggest that small localized changes within the community may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma. Both the drastically increased RR of CRC-associated genera versus diversity for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model. It is also possible that in a driver-passenger scenario, simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring. This would account for the observation that there is no constant additive effect on RR for increasing positivity. Additionally, the initial establishment of the driver within the system appears to be dependent on the current community. This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases.

### 2017-11-14 submission/manuscript.tex
add RR generation to methods section
add Will's comments discussion up to P2

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in individuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. Even though CRC associated genera
increases the relative risk of carcinoma it does not consistently
increase the relative risk of adenoma. This information suggests that
these specific genera are important in carcinoma genesis but may not be
the primary members of the microbial community that contribute to the
formation of an adenoma. Additionally, our data shows that by using the
whole community our models perform better then when they only use the
CRC associated genera. CRC associated genera are clearly important to
carcinoma but the context or community in which these microbes are a
part of can drastically increase the ability of models to make
predictions. This data supports the concept that small localized changes
within the community may be occuring that are important in the disease
progression of colorectal cancer and that they may not directly involve
CRC associated genera.

The driver-passenger model of the microbial role in CRC, as summarized
by Flynn {[}2{]}, can be supported with this data for carcinoma but not
necessarily for adenoma. The drasitically increased relative risk of
disease, when considering the CRC associated genera, is highly
supportive of this type of process, especially in the context of
increasing relative risk with more CRC associated genera positivity. It
is also possible that in a driver-passenger scenario that simply having
the driver present or only identifying the passenger is a good enough
proxy that the event is occuring. This would account for the observation
that there is no constant additive effect on relative risk for
increasing positivity. Additionally, the initial establishment of the
driver within the system is also dependent on the community that is
present and this is supported by the observation that when adding the
community context to our models along with the CRC associated genera the
model AUC increases.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
\(\alpha\)-diversity analysis, values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for \(\alpha\)-diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for study and variable region effects in feces and study,
variable region, and individual effects in tissue. Relative risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
Relative risk significance testing utilized the chi-squred test.
\(\beta\)-diversity differences utilized a Bray-Curtis distance matrix
and PERMANOVA executed with the vegan (v2.4.4) package. Random Forest
models were built using both the caret (v6.0.77) and randomForest
(v4.6.12) packages. Differences between the obtained AUC versus a random
model AUC was assessed using T-tests. Power analysis and estimations
were made using the pwr (v1.2.1) and statmod (v1.4.30) packages. All
figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
packages.

>>>>

Our study identifies clear differences in diversity both at the
community level and within individual genera that are present in
individuals with CRC versus those without the disease. Although there
was a step-wise decrease in diversity from control to adenoma to
carcinoma, this did not translate into large effect sizes for the
relative risk of either of these two conditions. Even though
CRC-associated genera increase the RR of carcinoma, they do not
consistently increase the relative risk of adenoma. This information
suggests that these specific genera are important in carcinoma genesis
but may not be the primary members of the microbial community
contributing to the formation of an adenoma. Additionally, our data show
that by using the whole community, our models perform better than when
only the CRC-associated genera are included. CRC-associated genera are
clearly important to carcinoma pathogenesis but accounting for the
community in which these microbes exist can drastically increase the
ability of models to make predictions. These observations suggest that
small localized changes within the community may be occuring that are
important in early disease progression of CRC and that this process may
not directly involve CRC-associated genera.

The data presented herein supports the driver-passenger model of the
microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
carcinoma but not necessarily adenoma. Both the drastically increased RR
of CRC-associated genera versus diversity for carcinoma and increasing
RR with more CRC-associated generea positivity are highly supportive of
this model. It is also possible that in a driver-passenger scenario,
simply having the driver present or only identifying the passenger is a
good enough proxy that the event is occuring. This would account for the
observation that there is no constant additive effect on RR for
increasing positivity. Additionally, the initial establishment of the
driver within the system appears to be dependent on the current
community. This is supported by our finding that when adding the
community context to our models in addition to the CRC-associated
genera, the model AUC increases.

@@

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
\(\alpha\)-diversity analysis, values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for \(\alpha\)-diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.14) package to
correct for study and variable region effects in feces and study,
variable region, and individual effects in tissue. Relative risk was
analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages by
assessing how many with and without disease were above and below the
overall median value within the specific study. Relative risk
significance testing utilized the chi-squred test. \(\beta\)-diversity
differences utilized a Bray-Curtis distance matrix and PERMANOVA
executed with the vegan (v2.4.4) package. Random Forest models were
built using both the caret (v6.0.77) and randomForest (v4.6.12)
packages. Differences between the obtained AUC versus a random model AUC
was assessed using T-tests. Power analysis and estimations were made
using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.
